
The response (binned Fisher) matrix $R_{AB}=\frac12\Tr[MP_AMP_B]$ is the
only piece of the QML pipeline whose cost grows faster than one solve per
data vector, so the engine choice decides feasibility at high resolution.
\simaster\ implements three engines (\S\ref{sec:fisher}); their error
budgets differ qualitatively:

\begin{itemize}
\item \textbf{Exact}: all $N_{\rm modes}$ columns of $MG$ solved;
  deterministic. Cost $N_{\rm modes}$ CG solves.
\item \textbf{Column-subsampled exact} (fraction $f$, stratified per
  $\ell'$, $N_{\ell'}/n_{\ell'}$-renormalized; exactly unbiased): the
  row index of $H$ is summed exactly, so the response error is local in
  bands, $\delta c_A/\sigma_A \simeq \mathrm{SNR}_A
  \sqrt{\rho(1-f)/n_A}$, with $\rho\simeq0.1$ measured for the
  NaMaster mask.
\item \textbf{Sims-MC} ($\mathrm{cov}[\hat y]$ over $N_{\rm sims}$
  fiducial sims): Wishart noise couples all bands,
  $\delta c_A/\sigma_A \simeq \sqrt{\mathrm{SNR}^2_{\rm tot}/
  N_{\rm sims}}$ with $\mathrm{SNR}^2_{\rm tot}=c^{\sf T}Rc$.
\end{itemize}

\paragraph{Requirement.} We demand that the Fisher-induced systematic not
dominate the error budget: offset $\le \epsilon\,\sigma_A$ per band with
$\epsilon=0.1$ (error-bar inflation $<0.5\%$); all counts below scale as
$1/\epsilon^2$.

\paragraph{Measured per-draw error budget.} The heuristic above estimates the
\emph{typical} subsampling offset; for a given run \simaster\ also
\emph{measures} it. With \texttt{run\_exact(sample\_frac}$=f$,
\texttt{keep\_samples=True)} the per-mode contributions to the binned $R$ and
noise bias $n$ are retained --- each solved column feeds exactly one column
band of the pre-symmetrized $R$ (a compact slab) and one band of $n$, since
the row index of $H$ is summed exactly --- and \texttt{subsample\_error()}
turns them into a covariance to \emph{add} to $R^{-1}$.  Both $R$ and $n$ are
linear in the per-mode contributions, so their subsampling covariance is the
exact stratified without-replacement estimator
\begin{equation}
  \mathrm{Cov}(\hat R\,c) \;=\; \sum_{\ell'} N_{\ell'}^2
  \Big(1-\frac{n_{\ell'}}{N_{\ell'}}\Big)\frac{S_{\ell'}}{n_{\ell'}},
  \label{eq:strat}
\end{equation}
with $S_{\ell'}$ the per-stratum sample covariance of the per-mode vectors and
the finite-population correction $(1-n_{\ell'}/N_{\ell'})$ --- the exact
per-$\ell'$ form of the $(1-f)$ factor --- making it vanish identically at
$f=1$.  Propagated by the delta method, $\delta c = -R^{-1}[\delta(Rc)+\delta
n]$: the $R$ and $n$ errors are \emph{correlated}, sharing the column draw, so
the honest bandpower budget combines them.  Because the matrix inverse is
nonlinear, a column \emph{bootstrap} is also provided --- resample the columns
within each stratum (FPC-scaled), rebuild $R^*$ and $n^*$ with the same
weights, and form $c^*=R^{*-1}(y-n^*)$ --- capturing the inverse's bias/skew
that the delta method misses, and yielding the otherwise-zero noise-bias error
$\sigma(n_A)$.  The diagnostic $\sqrt{\mathrm{diag}\,\mathrm{Cov}_{\rm
sub}/\mathrm{diag}\,R^{-1}}$ reports the per-band error-bar inflation directly
(a measured replacement for the $\epsilon$ requirement).  Single-mode strata
(the $\ge1$-column floor) contribute zero in both estimators, consistently ---
their within-stratum variance is undeterminable.  The per-mode store is
additive and concatenable, so a distributed run checkpoints it per node and
aggregates before this (cheap) post-processing.

\paragraph{Planck-like configuration.} $N_{\rm side}=1024$,
$\ell_{\max}=2000$, $T/E/B$, $f_{\rm sky}\simeq0.6$,
TT signal-dominated to $\ell\sim1500$, hence
$\mathrm{SNR}^2_{\rm tot}\approx10^6$ and
$N_{\rm modes}=3(\ell_{\max}+1)^2\approx1.2\times10^7$.
One $\C$-apply costs 30 core-s (measured on this machine's 24-core CPU (ducc0)); with
$\sim$entire-mission anisotropic noise we budget 100 CG iterations,
i.e.\ 0.84 core-h per solve.

\begin{table}[h]\centering\small
\begin{tabular}{lrr}
\toprule
engine & solves needed & CPU cost [core-h] \\
\midrule
exact (all columns) & $1.2\times10^7$ & 10\,088\,702 \\
sims-MC, no iteration & $\mathrm{SNR}^2_{\rm tot}/\epsilon^2
  = 10^8$ & 83\,988\,510 \\
subsampled, no iteration & $f\ge0.75$ of all columns &
  7\,566\,527 \\
\textbf{subsampled + iteration} & $\sim250$ cols/band
  $\to 49,500$\hspace{-2mm} & \textbf{41\,574} \\
sims-MC + iteration & $\sim39,600$ & 33\,259 \\
\bottomrule
\end{tabular}
\caption{Fisher cost for a Planck-like analysis at
$\epsilon=0.1$.  ``+ iteration'' means the \emph{around-fiducial}
estimator $\hat c = c_{\rm fid} + \hat R^{-1}(\hat y(d) -
\langle\hat y\rangle_{\rm fid\ sims})$ with one Newton--Raphson
re-centering of the fiducial: the sim mean carries the \emph{exact}
response of the filter, so the $\hat R$ error multiplies
$(c-c_{\rm fid})$ instead of $c$ (verified numerically: a 20\%-wrong
fiducial deflates the subsampled offsets by the predicted factor 5;
the plain $\hat R^{-1}(\hat y-\hat n)$ form does \emph{not}
benefit --- its bias is $R^{-1}\delta\hat R\,c$ for any fiducial).
The $1/\epsilon^2\approx100$ extra sims for
$\langle\hat y\rangle$ are negligible.  The budget is then set by the
\emph{covariance} accuracy of $R^{-1}$ ($\sim$2\%:
$n_A\ge\rho/2(0.01)^2\approx250$ columns per band, $\Delta\ell=30$,
3 components).}
\end{table}

Without iteration, neither stochastic engine is affordable for
signal-dominated data: the subsampling fraction obeys
$f/(1-f)\ge\rho f_{\rm sky}(S/(S+N))^2/2\epsilon^2$ --- note the band
width cancels --- giving $f\simeq0.75$, while sims-MC needs
$10^8$ solves.  With a good fiducial, the around-fiducial estimator and
one re-centering make the offset multiply $(c-c_{\rm fid})$ instead of
$c$, and both engines drop to
$\mathcal{O}(5\times10^4)$ solves: about 41\,574 core-h
($\approx$7 days on a 256-core
farm), or $\sim$1 week on a single A100-class GPU once a GPU SHT backend
exists ($\sim$0.15\,s per $\C$-apply), with $\sim$38\,GB of GPU RAM at
batch 64.  The noise bias never matters: Rademacher probes have
per-probe scatter equal to the per-realization $\sigma_A$, so
$1/\epsilon^2=100$ probes suffice.  We therefore consider the Planck
configuration feasible with the subsampled engine + iteration on a
mid-size CPU farm today, and on a single node with a GPU SHT.

\paragraph{Deflation gain.} The budget above charges 100 CG
iterations per solve, and that is exactly what deflated/recycled CG
(\S\ref{sec:deflation}) attacks.  Every solve of a given iteration shares
the same $\C$, so the slow-mode subspace is harvested once (a handful of
solves, amortized immediately) and projected out of all $\sim$49,500
solves; the deflation space is rebuilt only when the fiducial moves (twice,
negligible).  On the matrix-free \texttt{ducc} backend used here each CG
iteration is an SHT and the projector is SHT-free, so the measured
$1.5$--$2\times$ iteration reduction (\S\ref{sec:deflation}) carries
directly to wall-clock: the 41\,574 core-h headline drops to roughly
20\,787--27\,716
core-h ($\approx$3--5
days on the same 256-core farm), stacking multiplicatively with the
subsampling and around-fiducial savings already counted.  The reduction is
problem-dependent and tends to \emph{grow} with the condition number ---
i.e.\ with resolution and anisotropy --- for a fixed subspace size $k$, at
$O(N_{\rm row}\,k)$ memory; the figure here uses the conservatively
measured low-resolution value.
