\documentclass[11pt,a4paper]{article}
\usepackage[margin=22mm]{geometry}
\usepackage{amsmath,amssymb,graphicx,booktabs,hyperref,xcolor}
\graphicspath{{figs/}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Tr}{\mathrm{Tr}}
\newcommand{\simaster}{\textsc{SiMaster}}

\title{\simaster: GPU-accelerated quadratic maximum-likelihood power
  spectrum estimation\\[2mm]\large Implementation notes and validation report}
\author{generated with Claude Code for A.~Slosar}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
\simaster\ implements the optimal (quadratic maximum-likelihood, QML)
bandpower estimator for arbitrary collections of correlated spin-0 and
spin-2 fields on cut skies, with a NaMaster-like user interface.  No dense
pixel-covariance matrix is ever built or inverted: the inverse-covariance
filter is applied by preconditioned conjugate gradients whose operator costs
two spherical-harmonic transforms, batched over hundreds of right-hand sides
on a GPU.  The bandpower response (Fisher) matrix, noise bias and window
functions are obtained either \emph{exactly}, by batched-CG probing of all
harmonic modes (the default up to $N_{\rm side}\simeq 64$), or by a
Monte-Carlo engine that scales to $N_{\rm side}=1024$.  We validate the
estimator on three suites of 100 simulations at $N_{\rm side}=32$ (CMB
$T/Q/U$ with uniform and with strongly anisotropic noise, and galaxy
density $\times$ weak lensing with \texttt{pyccl} theory spectra), verify
$\chi^2$ distributions of the recovered bandpowers, and benchmark the cost
drivers to project the requirements of an $N_{\rm side}=1024$ analysis.
\end{abstract}

\tableofcontents

\section{Method}
\label{sec:method}

\subsection{Data model}
Each field provides three maps: \texttt{signal} (the observation),
\texttt{mask} ($w$, the multiplicative response), and \texttt{ivar} (the
per-pixel inverse noise variance), with
\begin{equation}
  d = w\,s + n,\qquad n\sim\mathcal N(0,\,\mathrm{ivar}^{-1})
  \quad\text{independently per pixel,}
\end{equation}
so the noise covariance $N$ is diagonal in pixel space by construction.
Pixels with $w=0$ or $\mathrm{ivar}=0$ are removed from the data vector.
Spin-2 fields contribute $(Q,U)$ maps and are estimated as $E/B$ spectra.

Internally all harmonic algebra uses the \emph{real} orthonormal spherical
harmonics: the coefficients of an isotropic field are i.i.d.\ with variance
$C_\ell$, the per-mode spectral matrices $(N_c\times N_c)$ are real, and all
operators are real GEMMs.  The concatenated data vector has covariance
\begin{equation}
  \C \;=\; U\hat C U^{\sf T} + N \;+\;\textstyle\sum_j \alpha_j t_j t_j^{\sf T},
  \qquad U = W\,Y\,B,
  \label{eq:cov}
\end{equation}
with $Y$ the real-basis synthesis matrix restricted to observed pixels, $W$
the mask, $B$ optional beams, $\hat C$ the block-diagonal fiducial spectra
(one $N_c\times N_c$ block per mode), and $t_j$ templates to marginalize.

\subsection{Estimator}
With $A=(\text{spectrum pair},\text{band})$ and $P_A=\partial\C/\partial c_A$
(flat in $C_\ell$ within each band),
\begin{equation}
  \hat y_A = \tfrac12\, d^{\sf T} M P_A M d,\qquad
  \hat c = R^{-1}\big(\hat y - n\big),\qquad
  n_A = \tfrac12\Tr[M N M P_A],\qquad
  R_{AB} = \tfrac12\Tr[M P_A M P_B],
  \label{eq:qml}
\end{equation}
where $M$ is the inverse-covariance filter ($\C^{-1}$, or the
template-deprojected pseudo-inverse of \S\ref{sec:templates}).  The
estimator is unbiased for \emph{any} fiducial spectrum --- only optimality
degrades --- which justifies single-shot estimation around a fiducial;
\texttt{iterate()} re-centers the fiducial on the estimate (one
Newton--Raphson step on the field-level likelihood per iteration).
$\mathrm{cov}(\hat c)\simeq R^{-1}$ at the fiducial.

Evaluating $\hat y$ costs one CG solve plus one adjoint SHT: with
$a = B^{\sf T}Y^{\sf T}W^{\sf T} (Md)$, each band statistic is
$\hat y_{(cd),b} = (1+\delta_{cd})^{-1}\sum_{k\in b} (1+\delta_{cd}\!-\!\delta_{cd}/2)\dots$
concretely $\tfrac12\sum_k a_{c,k} a_{c,k}$ for autos and
$\sum_k a_{c,k} a_{d,k}$ for crosses.

\subsection{Inverse-covariance filter: CG with a provably SPD preconditioner}
\label{sec:cg}
$\C\,x$ costs two SHTs plus diagonal work and is batched over many
right-hand sides.  The preconditioner is the Woodbury inverse of the
isotropic approximation to (\ref{eq:cov}),
\begin{equation}
  P^{-1} = N^{-1} - N^{-1} U\, T\, U^{\sf T} N^{-1},\qquad
  T_\ell = \big(\hat C_\ell^{-1} + D_\ell\big)^{-1}
  = \hat C - \hat C D^{1/2}\big(1+D^{1/2}\hat C D^{1/2}\big)^{-1}D^{1/2}\hat C,
\end{equation}
the right-hand form remaining valid for singular $\hat C_\ell$ (e.g.\ zero
fiducial $BB$).  Positive definiteness of $P^{-1}$ requires $D$ to dominate
the spectrum of $U^{\sf T}N^{-1}U$; the guaranteed bound is
$2\max_p(w^2\,\mathrm{ivar})\,N_{\rm pix}/4\pi$ (the factor 2 covers the
HEALPix Gram-matrix eigenvalue spread, measured to reach 1.75 at
$\ell=3N_{\rm side}$).  Because the bound is loose for strongly varying
$w^2\,\mathrm{ivar}$, \simaster\ starts from the statistical mean
$\sum_p w^2\,\mathrm{ivar}/4\pi$ and escalates toward the bound only if CG
detects an indefinite direction (monitored every iteration at no cost).
Uniform-weight problems converge in $\sim$10 iterations at
$10^{-5}$ tolerance; the factor-2 anisotropic-noise validation costs
$\sim$2--3$\times$ more.

\subsection{Response matrix, noise bias, windows}
\label{sec:fisher}
Three engines, validated against each other and against a brute-force dense
reference at small $N_{\rm side}$ (the third, column-subsampled exact
engine, is described with measurements in \S\ref{sec:validation} and
costed in \S\ref{sec:fishercost}):

\paragraph{Exact (default for $N_{\rm side}\lesssim 64$).}  Solving
$V = M G$ for all $N_{\rm modes}$ columns of $G=U$ (batched CG; never
storing the dense $H = G^{\sf T} M G$) and accumulating the $\ell$-resolved
response tensor
\begin{equation}
  T_{abcd}(\ell,\ell') \;=\; \sum_{k\in\ell,\,k'\in\ell'}
  H_{ak,ck'}\,H_{bk,dk'}
\end{equation}
chunk by chunk yields the exact $R$, the exact per-multipole window
functions $F_{b\ell}$, and (from the diagonal blocks of $V^{\sf T}NV$) the
exact noise bias, for any binning.  Cost equals $N_{\rm modes}$ CG
right-hand sides, i.e.\ comparable to a few-thousand-simulation MC run, but
deterministic.

\paragraph{Monte Carlo (any $N_{\rm side}$).}  For sims
$x\sim\mathcal N(0,\C)$, $\mathrm{cov}[\hat y_A,\hat y_B] = R_{AB}$
exactly (also for the deprojected filter, since $M\C M = M$); the same sims
give $F_{b\ell}=\mathrm{cov}[\hat y_b, \hat y_\ell]$ and noise-only sims
give $n$.  \textbf{Caveat (important):} the frozen MC noise of $\hat R$
imprints a fixed offset on all estimates with r.m.s.
\begin{equation}
  \sigma_{\rm off}(\hat c_A) \;\approx\; \sigma_A\,
  \sqrt{\mathrm{SNR}^2_{\rm tot}/N_{\rm sims}},
  \qquad \mathrm{SNR}^2_{\rm tot} = c^{\sf T} R\, c,
\end{equation}
i.e.\ it scales with the \emph{total} squared signal-to-noise across all
bands ($\sim$ the number of signal-dominated modes).  For signal-dominated
data this demands large $N_{\rm sims}$; the exact engine avoids the issue
entirely and is what the validation below uses.  The Hartlap factor is
applied to $\hat R^{-1}$.

\subsection{Bands, junk bands, aliasing}
Bandpowers are flat in $C_\ell$.  User bands are augmented with ``junk''
bands so the band basis covers all of $[\ell_{\min},\ell_{\max}]$; all bands
are estimated jointly and junk bands are marginalized via the full
$R^{-1}$.  The covariance model is bandlimited at $\ell_{\max}$ (default
$3N_{\rm side}-1$): data power above $\ell_{\max}$ is not modeled and would
alias into the estimates --- the resolution/$\ell_{\max}$ must be chosen so
that signal+noise above the limit is negligible.  Validation simulations are
generated bandlimited, making the model exact.  Monopole and dipole of
spin-0 fields lie outside the band basis and are deprojected automatically.

\subsection{Template marginalization}
\label{sec:templates}
Two prescriptions, equivalent and validated against each other:
(i) \texttt{template\_alpha}$\,=\alpha$ adds
$\alpha\,\mathrm{tr}(\C)/\lVert t\rVert^2\; t t^{\sf T}$ to the covariance
(the classic ``correctly large prefactor''; also included when drawing MC
sims so that $R$ stays consistent; $\alpha\lesssim10^6$ is safe in double
precision); (ii) the default $\alpha\to\infty$ limit via
Sherman--Morrison--Woodbury,
$M = \C^{-1} - \C^{-1}T\,(T^{\sf T}\C^{-1}T)^{-1}T^{\sf T}\C^{-1}$,
implemented as one extra batched solve, numerically stable and exactly
nulling the template directions.

\section{Implementation}
\label{sec:impl}
Pure Python.  All solver linear algebra runs in JAX (float64) on GPU or
CPU transparently.  Two exact SHT backends:
\begin{itemize}
\item \texttt{dense}: the real-basis synthesis matrices restricted to
  observed pixels are precomputed (threaded \texttt{ducc0} syntheses, disk
  cache) and live on the GPU; every SHT is a GEMM and the batched CG is
  GEMM-bound --- ideal up to $N_{\rm side}\simeq64$ on a small GPU.
\item \texttt{ducc}: matrix-free transforms through \texttt{ducc0} (the
  engine behind healpy; exact adjoint pairs, multithreaded CPU) wrapped in
  \texttt{jax.pure\_callback}; memory $O(N_{\rm obs})$, scales to
  $N_{\rm side}=1024$ and beyond.
\end{itemize}
\texttt{s2fft} (native-JAX GPU SHTs) was evaluated and rejected: its
HEALPix spin-2 synthesis carries 5--13\% pointwise errors in v1.4.0
(spin-0 is exact to $10^{-14}$), fatal for QML; the backend interface is
ready for a future GPU SHT.

The NaMaster-style interface (\texttt{Field}, \texttt{Bins},
\texttt{compute\_full\_master}) is described in \texttt{docs/migration.md}.

\section{Validation at $N_{\rm side}=32$}
\label{sec:validation}
\input{validation_section}

\section{Performance and $N_{\rm side}=1024$ feasibility}
\label{sec:feasibility}
\input{feasibility_section}

\section{Fisher-matrix methods and the cost of a Planck-like analysis}
\label{sec:fishercost}
\input{fisher_methods_section}

\section{Numeric caveats (collected)}
\begin{enumerate}
\item Everything is float64; float32 is not supported (quadratic
  estimators difference large numbers).
\item CG tolerance ($10^{-5}$ default) enters data and response through the
  same filter, so the leading effect cancels in $R^{-1}\hat y$; do not
  relax beyond $\sim10^{-4}$.
\item MC-mode frozen-$\hat R$ offsets scale with total S/N
  (\S\ref{sec:fisher}); the exact engine is preferred whenever affordable.
\item The bandlimit $\ell_{\max}$ caveat (\S\ref{sec:method}): unmodeled
  high-$\ell$ power aliases in.  HEALPix quadrature non-orthonormality at
  $\ell>2N_{\rm side}$ is part of the model (not an error), but inflates
  the variance of the top bands --- treat bands near $3N_{\rm side}$ as junk.
\item With curved input spectra, compare to the window-convolved prediction
  (\texttt{predict()}), not to naively binned theory.
\item The preconditioner escalation (\S\ref{sec:cg}) trades iterations for
  safety under strongly anisotropic weights.
\item \simaster\ is curved-sky only (no flat-sky estimator).  The only
  geometry-aware object is the synthesis operator $Y$, so a flat-sky port
  (FFTs, $|k|$ bins) would reuse the solver, response engines and
  compression unchanged --- future work.
\end{enumerate}

\end{document}
